Body Mass Index and Pathway Dysregulation in Cancer

### 2016-09-12 presentation/beamer_theme/images/tmp/pdf/colloquia/beamer_theme/example.tex
draft ctcr presentation

\begin{frame}{Overview}
	\begin{enumerate}
		\item Introduction to obesity and cancer
		\item Results from last committee meeting
		\item Results since the last committee meeting
		\item Future directions
	\end{enumerate}
\end{frame}

@@

\begin{frame}
	\frametitle{Introduction}
	\begin{center}
		\includegraphics[width=0.48\paperwidth]{./images/ob1999}\\
		1999\\
		\includegraphics[width=0.48\paperwidth]{./images/ob2014}\\
		2014
	\end{center}
\end{frame}

@@

\begin{frame}
	\frametitle{Obesity-associated Genetic Signatures}
	Creighton \textit{et al.} (2012)
	\begin{itemize}
		\item Identified obesity-associated genetic signatures from the tumours of breast cancer patients
		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
	\end{itemize}
	Fuentes-Mattei \textit{et al.} (2014)
	\begin{itemize}
		\item Identified obesity-associated genetic signatures from the tumours of ER+ breast cancer patients
		\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
	\end{itemize}
\end{frame}

@@

\begin{frame}{Progress Overview}
	Where I was at the last committee meeting:
	{\footnotesize
		\begin{itemize}
			\item Applied Creighton metagene to ICGC data
			\item Applied Fuentes-Mattei metagene to Creighton \textit{et al.} and ICGC data
			\item Found no association with BMI status
		\end{itemize}
	}
	\vskip2ex
	What I have done since the last committee meeting:
	{\footnotesize
		\begin{itemize}
			\item Gene expression analysis in Creighton \textit{et al.} data
			\item Gene expression analysis in ICGC data
			\item Pathway enrichment analysis in ICGC data
			\item Comparison of the obesity metagenes with other pathway metagenes from Gatza \textit{et al.} (preliminary results)
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Clinical Data with Patient BMI}
	Out of 33 cancer types, 14 types had both height and weight data\\
	{\footnotesize
		\begin{itemize}
			\item 8 cancer types had RNA-seq data\\
		\end{itemize}
	}
	\vskip2ex
	\begin{center}
		\begin{tabular}{lcccl}
			& Normal & Overweight & Obese & Total\\
			\rowcolor{LightCyan}
			BLCA & 103  & 97         & 61    & 261\\
			CESC & 85   & 62         & 77    & 224   \\
			\rowcolor{LightCyan}
			COAD & 75   & 80         & 71    & 226   \\
			KIRP & 26   & 53         & 45    & 124\\
			\rowcolor{LightCyan}
			LIHC & 140  & 75         & 49    & 264\\
			READ & 22   & 36         & 15    & 73\\
			\rowcolor{LightCyan}
			SKCM & 72   & 73         & 73    & 218\\
			UCEC & 90   & 106        & 286   & 482\\
			\hline
			\rule{0pt}{2.25ex}Total & 613 & 582 & 677 & 1872\\
		\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{Validation of Methodology -- Creighton \textit{et al.} (2012)}
	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
	{\footnotesize
		\begin{itemize}
			\item Majority of these genes were downregulated in obese patients
		\end{itemize}
	}
	\begin{center}
		\includegraphics[width=0.6\paperwidth]{./images/creighton}
	\end{center}
\end{frame}

@@

\begin{frame}{Summary 1}
	Metagene from Creighton \textit{et al.} correlated with gene expression and patient BMI in Creighton data set
	\vskip2ex
	Metagene derived from Creighton's obesity-associated genes  correlated with the gene expression but not the patient BMI in ICGC data
	\vskip2ex
	Metagene derived from Fuentes-Mattei obesity-associated genes correlated with the gene expression but not the patient BMI in both Creighton \textit{et al.} and ICGC data
\end{frame}

\begin{frame}{Gene Expression Analysis}
	Maybe the obesity-associated genes identified by either group were not related to obesity, but related to a different clinical variable
	\vskip2ex
	Are the differentially expressed genes between the obese and non-obese groups (after correcting for other clinical variables) similar to the obesity-associated genes?
	{\footnotesize
		\begin{itemize}
			\item How do these DEGs compare with the Creighton's or Fuentes-Mattei's obesity-associated genes?
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Residual data}
	All of the non-BMI clinical variables were controlled for in the linear model
	{\footnotesize
		\begin{itemize}
			\item Age, ethnicity, menopause status, grade, LN status, ER/PR/HER2 statuses
		\end{itemize}
	}
	\vskip2ex
	Residual data (or remaining data after controlling for the variables) was used to find the DEGs
	\vskip2ex
	Again, more than 799 gene probes showed up (1104 genes)
	{\footnotesize
		\begin{itemize}
			\item 168 genes from Creighton \textit{et al.} were found in the top 799 genes
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Completely remove the effect of ethnicity}
	African American patients are more likely to have higher BMI than Cauacasian patients
	\vskip2ex
	Removed the effect of ethnicity by removing the African American patients' data (total of 77 patient data to work with)
	{\footnotesize
		\begin{itemize}
			\item All of the other clinical variables were adjusted as before
		\end{itemize}
	}
	\vskip2ex
	Top 799 genes were identified in this data set (both raw and residual)
	{\footnotesize
		\begin{itemize}
			\item  148 and 92 genes were in the top 799 genes in the raw and residual data, respectively
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Metagene Analysis}
	All of the gene probe sets I have found (8 in total) were used to create metagenes and to transform other cancer types
	\begin{itemize}
		\item Checked in Creighton \textit{et al.} data
		\item Transferred into other cancer data sets
	\end{itemize}
\end{frame}

@@

\begin{frame}{Do the metagenes correlate with GE in ICGC data?}
	\includegraphics[width=0.23\paperwidth]{./images/rawobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/resobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/caobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/caresobsmeta2}\\
	\includegraphics[width=0.23\paperwidth]{./images/crolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/rescrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/cacrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/carescrolmeta2}
\end{frame}

\begin{frame}{Do the metagenes correlate with BMI?}
	\includegraphics[width=0.23\paperwidth]{./images/rawobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/resobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/caobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/caresobsclin3}\\
	\includegraphics[width=0.23\paperwidth]{./images/crolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/rescrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/cacrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/carescrolclin3}
\end{frame}

@@

\begin{frame}{Metagenes are correlating with GE but not BMI}
	All of the metagenes found so far have correlated with the overall gene expression of the samples, but not with the sample BMI
	\vskip2ex
	The metagenes are picking up gene signatures  that are common across all cancer types, but not related to BMI
	\vskip2ex
	\begin{center}
		\textbf{So what signature/mechanism is the metagene picking up?}
	\end{center}
\end{frame}

@@

\begin{frame}{Results}
	\begin{center}
		\includegraphics[width=0.90\paperwidth]{./images/allmeta1}
	\end{center}
\end{frame}

\begin{frame}{Results}
	\begin{center}
		\includegraphics[width=0.90\paperwidth]{./images/allmeta2}
	\end{center}
\end{frame}

\begin{frame}{Results}
	\begin{center}
		\includegraphics[width=0.45\paperwidth]{./images/allmetacor1}
	\end{center}
\end{frame}

\begin{frame}{Summary 2}
	Metagenes found in Creighton \textit{et al.} data correlates with GE and BMI status in Creighton data, but not in  the ICGC data
	\vskip2ex
	The metagenes do not correlate with any other pathway metagenes
	\vskip2ex
	\begin{center}
		\textbf{Are there any common genes and/or pathways being dysregulated across all of the cancer types?}
	\end{center}
\end{frame}

@@

\begin{frame}{Pathway Enrichment Analysis}
	Maybe each cancer type have a different set of genes activated
	\vskip2ex
	Are there any specific pathways enriched in any specific cancer type?
	\vskip2ex
	If so, is it common across other cancer types?
	\vskip2ex
	How relevant is the pathway to obesity and/or cancer?
	\vskip2ex
	Can we use it to narrow down our search for obesity-associated signature?
\end{frame}

@@

\begin{frame}{Discussion}
	Different methods used to find DEGs
	{\footnotesize
		\begin{itemize}
			\item Found more genes than Creighton \textit{et al.} because they were using different programs and normalisation method in their analyses
		\end{itemize}
	}
	\vskip2ex
	Tumour heterogeneity within and between samples
	{\footnotesize
		\begin{itemize}
			\item Not finding any significant genes or pathways because the samples are so different from one another
		\end{itemize}
	}
	\vskip2ex
	Tumour sample quality and size
	{\footnotesize
		\begin{itemize}
			\item Picking up too much noise and/or don't have enough samples to find anything significant
		\end{itemize}
	}
\end{frame}

\begin{frame}{Discussion -- Inter-tumour heterogeneity 1}
	Each patient has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
	\vskip2ex
	Each data set will have a different collection of patients, which in turn will have a different collection of dysregulated pathways
	{\footnotesize
		\begin{itemize}
			\item Hence DEGs show up, but not the pathways
		\end{itemize}
	}
\end{frame}

\begin{frame}{Discussion -- Inter-tumour heterogeneity 2}
	Some of the high BMI patients in the data set will have tumours that are BMI driven, but BMI may not be the driver for some of these patients
	\begin{itemize}
		\item Only subset of the patients with high BMI are actually driven by BMI
	\end{itemize}
	\vskip2ex
	Other data sets may have less of these truly BMI driven tumours
	\begin{itemize}
		\item Therefore no association with the signatures
	\end{itemize}
\end{frame}

@@

\begin{frame}{test}
	\citet{Fuentes-Mattei2014}\\
	\citet{Creighton2012}\\
	\citet{Alexandrov2013}
	\cite{Gatza2010a}
\end{frame}

>>>>

@@

\begin{frame}
\frametitle{Introduction}
\begin{center}
	\includegraphics[width=0.48\paperwidth]{./images/ob1999}\\
	1993\\
	\includegraphics[width=0.48\paperwidth]{./images/ob2014}\\
	2014
\end{center}
\end{frame}

@@

\begin{frame}
	\frametitle{Obesity-associated Genetic Signatures}
	Creighton \textit{et al.} (2012)
	\begin{itemize}
		\item Identified obesity-associated genetic signatures in breast cancer patients
		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
	\end{itemize}
	Fuentes-Mattei \textit{et al.} (2014)
	\begin{itemize}
		\item Identified obesity-associated genetic signatures in ER+ breast cancer patients
		\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
	\end{itemize}
\end{frame}

@@

\begin{frame}{Progress Overview}
	Where I was at the last committee meeting:
	{\footnotesize
		\begin{itemize}
			\item Applied Creighton metagene to ICGC data
			\item Applied Fuentes-Mattei metagene to Creighton \textit{et al.} and ICGC data
			\item Found no association with BMI status
		\end{itemize}
	}
	\vskip2ex
	What I have done since the last committee meeting:
	{\footnotesize
		\begin{itemize}
			\item Gene expression analysis using Creighton \textit{et al.} data
			\item Gene expression analysis using ICGC data
			\item Pathway enrichment analysis using ICGC data
			\item Comparison of the obesity metagenes with other pathway metagenes from Gatza \textit{et al.} (preliminary results)
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Clinical Data with Patient BMI}
	Out of 33 cancer types, 14 types had both height and weight data\\
	{\footnotesize
		\begin{itemize}
			\item 8 cancer types had RNA-seq data\\
		\end{itemize}
	}
	\vskip2ex
	\begin{center}
		\begin{tabular}{lcccl}
			& Lean & Overweight & Obese & total \\
			\rowcolor{LightCyan}
			BLCA & 103  & 97         & 61    & 261\\
			CESC & 85   & 62         & 77    & 224   \\
			\rowcolor{LightCyan}
			COAD & 75   & 80         & 71    & 226   \\
			KIRP & 26   & 53         & 45    & 124\\
			\rowcolor{LightCyan}
			LIHC & 140  & 75         & 49    & 264\\
			READ & 22   & 36         & 15    & 73\\
			\rowcolor{LightCyan}
			SKCM & 72   & 73         & 73    & 218\\
			UCEC & 90   & 106        & 286   & 482\\
			\hline
			\rule{0pt}{2.25ex}Total & 613 & 582 & 677 & 1872\\
		\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{Validation of Methodology -- Creighton \textit{et al.} (2012)}
	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
	{\footnotesize
		\begin{itemize}
			\item These gene probes were used to assign an obesity-associated gene signature score to each sample
		\end{itemize}
	}
	\begin{center}
		\includegraphics[width=0.6\paperwidth]{./images/creighton}
	\end{center}
\end{frame}

@@

\begin{frame}{Summary 1}
	Metagene from Creighton \textit{et al.} correlated with gene expression and patient BMI in Creighton data set
	\vskip2ex
	Metagene from Creighton \textit{et al.} correlated with gene expression but not the patient BMI in ICGC data
	\vskip2ex
	Metagene from Fuentes-Mattei \textit{et al.} correlated with gene expression but not the patient BMI in both Creighton \textit{et al.} and ICGC data
\end{frame}

\begin{frame}{Gene Expression Analysis}
	Maybe the obesity-associated genes identified by either group were not related to obesity, but related to a different clinical variable
	\vskip2ex
	Does the differentially expressed genes between the obese and non-obese groups (after correcting for other clinical variables) similar to the obesity-associated genes?
	{\footnotesize
		\begin{itemize}
				\item How do these DEGs compare with the Creighton's or Fuentes-Mattei's obesity-associated genes?
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Residual data}
	All of the non-BMI clinical variables were controlled for in the linear model
	{\footnotesize
		\begin{itemize}
			\item Age, ethnicity, menopause status, grade, LN/ER/PR/HER2 status
		\end{itemize}
	}
	\vskip2ex
	Residual data (or remaining data after controlling for the variables) was used to find the DEGs
	\vskip2ex
	Again, more than 799 gene probes showed up (1104 genes)
	{\footnotesize
		\begin{itemize}
			\item 168 genes from Creighton \textit{et al.} were found in the top 799 genes
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Completely remove the effect of ethnicity}
	African American patients are more likely to be obese than Cauacasian patients
	\vskip2ex
	Removed the effect of ethnicity by removing the African American patients' data (total of 77 patient data to work with)
	{\footnotesize
		\begin{itemize}
			\item All of the other clinical variables were adjusted as before
		\end{itemize}
	}
	\vskip2ex
	Top 799 genes were identified in this data set (both raw and residual)
	{\footnotesize
		\begin{itemize}
			\item  148 and 92 genes were in the top 799 genes in the raw and residual data, respectively
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{Metagene Analysis}
	All of the new/novel gene probe sets (8 in total) were used to create metagenes and used to transform other cancer types
	\begin{itemize}
		\item Checked in Creighton \textit{et al.} data
		\item Transferred into other cancer data sets
	\end{itemize}
\end{frame}

@@

\begin{frame}{Do the metagenes correlate with gene expression in ICGC data?}
	\includegraphics[width=0.23\paperwidth]{./images/rawobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/resobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/caobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/caresobsmeta2}\\
	\includegraphics[width=0.23\paperwidth]{./images/rawcrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/rescrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/cacrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/carescrolmeta2}
\end{frame}

\begin{frame}{Do the metagenes correlate with BMI?}
	\includegraphics[width=0.23\paperwidth]{./images/rawobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/resobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/caobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/caresobsclin3}\\
	\includegraphics[width=0.23\paperwidth]{./images/rawcrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/rescrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/cacrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/carescrolclin3}
\end{frame}

@@

\begin{frame}{Metagenes are correlating with GE but not BMI}
	All of the metagenes found so far have correlated with the overall gene expression of the samples, but not with the sample BMI
	\vskip2ex
	The metagenes are picking up gene signatures  that are common across all cancer types, and not related to BMI
	\vskip2ex
	\textbf{So what signature/mechanism is the metagene picking up?}
\end{frame}

@@

\begin{frame}{Results}
	\begin{center}
		\includegraphics[width=0.90\paperwidth, height=0.60\paperheight]{./images/allmeta1}
	\end{center}
\end{frame}

\begin{frame}{Results}
	\begin{center}
		\includegraphics[width=0.50\paperwidth]{./images/allmeta2}
	\end{center}
\end{frame}

\begin{frame}{Summary 2}
	Metagenes found in Creighton \textit{et al.} data correlates with GE and BMI status in Creighton data, but not in  the ICGC data
	\vskip2ex
	The metagenes do not correlate with any other pathway metagenes
	\vskip2ex
	\textbf{Are there any common genes and/or pathways being dysregulated across all of the cancer types?}
\end{frame}

@@

\begin{frame}{Pathway Enrichment Analysis}
	Maybe each cancer type have a different set of genes activated
	\vskip2ex
	Are there any pathways enriched by these genes?
	\vskip2ex
	If so, is it common across other cancer types?
	\vskip2ex
	How relevant is the pathway to obesity and/or cancer?
	\vskip2ex
	Can we use it to narrow down our search for obesity-associated signature?
\end{frame}

@@

\begin{frame}{Discussion}
	Different methods used to find DEGs
	{\footnotesize
		\begin{itemize}
			\item Found more genes than Creighton \textit{et al.} because they were using different programs and normalisation method in their analyses
				\begin{itemize}
					\item Logged the data instead of RMA normalisation
				\end{itemize}
		\end{itemize}
	}
	\vskip2ex
	Tumour heterogeneity within and between samples
	{\footnotesize
		\begin{itemize}
			\item Not finding any significant genes or pathways because the samples are so different from one another
		\end{itemize}
	}
	\vskip2ex
	Tumour sample quality and size
	{\footnotesize
		\begin{itemize}
			\item Picking up too much noise and/or don't have enough samples to find anything significant
		\end{itemize}
	}
\end{frame}

\begin{frame}{Discussion -- Tumour heterogeneity}
	Each sample has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
	{\footnotesize
		\begin{itemize}
			\item Each data set will have more/less of these samples
			\item Therefore picking up DEGs that are significant in that one particular data set but not the other data sets
		\end{itemize}
	}
	\vskip2ex
	Each data set will have a different collection of samples, which in turn have a different collection of dysregulated pathways
	{\footnotesize
		\begin{itemize}
			\item Hence DEGs show up, but not the pathways
		\end{itemize}
	}
\end{frame}

@@

\begin{frame}{test}
	\citet{Fuentes-Mattei2014}\\
	\citet{Creighton2012}\\
	\citet{Alexandrov2013}
	\citet{Gatza2010}
\end{frame}
